---
title: "Bipin"
output: html_document
date: "2022-11-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
footballgames <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-04/games.csv')

##### derive categorical variable #####
# replace NA's in tie column with string 0
footballgames$tie[is.na(footballgames$tie)] <- "0"
# replace ties with string 1 instead of string with team name
for(i in 1:5324){
  if(footballgames$tie[i] != "0"){
    footballgames$tie[i] <- "1"
    i < i+1
  }
  else{
    i <- i+1
  }
}

# treat tie as categorical
footballgames$tie <- as.factor(footballgames$tie)


# get structure to check if tie is not a factor
str(footballgames)

```

```{r, echo = TRUE, results = 'hide'}
# replace ties with string 1 instead of string with team name
for(i in 1:5324){
  if(footballgames$tie[i] == 1){
    footballgames$winner[i] <- "Tie"
    i < i+1
  }
  else{
    i <- i+1
  }
}

footballgames
```

#### AFC or NFC, who is winning the most?

The National Football League (NFL) is a composition of two conferences AFC (American Football Conference) and NFC (National Football Conference). Here, the objective is to find out who have won most games (Home and Away) whenever an AFC team meets with a NFC team. Additionally, to see the distribution of the points won by each conference. In order to do this, two new columns 'HT' and 'AT' is created to find out if the home and away playing team was AFC or NFC with reference from the list of AFC teams and NFC teams and the variables 'home_team' and 'away_team'. However, there were games that was played between two AFC teams as well as two NFC teams which we are not interested in. Also, the Superbowl games are played in the neutral venue, so it would not make sense to have them included here. Then,the data was filter out for only the matches between AFC and NFC teams. There are 5324 matches in the dataset. Out of 5324 games, only 1288 matches were played between an AFC team and NFC team either home or away. The respective number of wins for each conference is then calculated and a bar plot of the same is created.Further, a box plot of distribution of points won for each conference is created.

```{r, echo = FALSE, warning=FALSE, message = FALSE}
####Calculating the percentages of AFC and NFC wins whenever they face each other.
library(tidyverse)
library(magrittr)
library(dplyr)
library(ggplot2)
library(plotly)
library(dplyr)

#AFC vs NFC teams list
afc_teams<- c("Cleveland Browns","Baltimore Ravens","Pittsburgh Steelers","Cincinnati Bengals","Miami Dolphins","Buffalo Bills","New York Jets","New England Patriots", "Kansas City Chiefs","Los Angeles Chargers","Denver Broncos","Las Vegas Raiders","Tennessee Titans","Houston Texans","Indianapolis Colts","Jacksonville Jaguars", "Oakland Raiders", "San Diego Chargers")

nfc_teams<-c("Minnesota Vikings","Chicago Bears","Detroit Lions","Green Bay Packers","New York Giants","Dallas Cowboys","Washington Redskins","Philadelphia Eagles","Arizona Cardinals","Los Angeles Rams","Seattlle Seahawks","San Francisco 49ers","Atlanta Falcons","Tampa Bay Buccaneers","Carolina Panthers ","New Orleans Saints", "St. Louis Rams")

```

```{r,echo = FALSE, warning=FALSE, message = FALSE}
#Create a new column 'HT' to find which was the home team: AFC or NFC

for (i in 1:5324){
  if(footballgames$home_team[i] %in% afc_teams){
    footballgames$HT[i]<- "AFC"
  }
  else{
    footballgames$HT[i]<-"NFC"
  }
}



#Create a new column 'AT' to find which was the away team: AFC or NFC
for (i in 1:5324){
  if(footballgames$away_team[i] %in% afc_teams){
    footballgames$AT[i]<- "AFC"
  }
  else{
    footballgames$AT[i]<-"NFC"
  }
}

#Find the games AFC team vs NFC team, home and away. 

afcvsnfc<- filter(footballgames, week!= "SuperBowl", HT=="AFC" & AT=="NFC"| HT=="NFC" & AT=="AFC")
dim(afcvsnfc)
#Find out who won against AFC vs NFC only, home and away. 
for(i in 1:1288){
  if(afcvsnfc$winner[i]=="Tie"){
    afcvsnfc$win_conf[i]="Tied"
    }
    else if(afcvsnfc$winner[i] %in% afc_teams){
    afcvsnfc$win_conf[i]<- "AFC"
    }
    else{
    afcvsnfc$win_conf[i]<- "NFC"
    }  
}

#Calculate the percentage of games won by NFC and AFC teams.
table(afcvsnfc$win_conf)
#Whenever, AFC played NFC either home or away:
#AFC won 661 of the games vs. NFC 623. Remaining 4 matches ended in a tie.
```

```{r, echo = FALSE, warning=FALSE, message = FALSE}
#Barplot of number of games won, AFC vs NFC playing each other only. 
plot_afcvsnfc_total_win<-afcvsnfc %>%
  ggplot(aes(x = win_conf, fill = win_conf)) + 
  geom_bar() + 
  geom_text(stat='count', aes(label=..count..), vjust=0) + 
  ggtitle("Number of games won by respective divison playing home and away ") +
  xlab("Winning Division Team") +ylab("Number of games won")+guides(fill=guide_legend(title="Win Conf."))
    scale_fill_discrete(labels=c('AFC Win', 'NFC Win','Tied')) 

plot_afcvsnfc_total_win 
```

```{r,echo = FALSE, warning=FALSE, message = FALSE}
pts_afcvsnfc<- ggplot(filter(afcvsnfc, winner!= "Tie"))+geom_boxplot(aes(x=win_conf,y=pts_win, fill= win_conf))+xlab("Winning Conference")+ylab("Points Won")+scale_y_continuous(limits = c(0, 60))+ggtitle("Points won by AFC Vs NFC Conference teams at home and away.")+ guides(fill=guide_legend(title="Win Conf."))

pts_afcvsnfc
  
```

#### What does the number of wins at home looks like for AFC vs NFC teams?

Here, we want to see who wins most games at home, NFC's teams or AFC's teams. In order to achieve this, first NFC teams home games are filtered out from the original data set. It is to be noted that Superbowls games are not included, since they are played in a neutral venue and matches between two teams from same conference is also excluded. In this new dataset, a new column called 'win_conf' is created that stores the result of the winning team's divison. Then, the number of games won by each division teams whenever NFC played at home is calculated. Similar steps are carried out whenever the AFC teams are playing at home. Finally, a side-by-side bar-plots are created for comparison.

```{r,echo = FALSE, warning = FALSE, message = FALSE}
library(grid)
library(gridExtra)
#Number of games won at home by NFC teams whenever they played AFC teams.

###First create a data with home team being NFC team and away team being AFC team. 
homewin_nfc<-filter(footballgames, week!= "SuperBowl", footballgames$home_team %in% nfc_teams & footballgames$away_team %in% afc_teams )

dim(homewin_nfc)

###To find out which division won when NFC played at home. 
for(i in 1:557){
  if(homewin_nfc$winner[i]=="Tie"){
     footballgames$win_conf[i]<-"Tied"
  }
  else if(homewin_nfc$winner[i]%in% afc_teams){
    homewin_nfc$win_conf[i]<-"AFC_Win"
  }
  else{
    homewin_nfc$win_conf[i]<-"NFC_Win"  
  }
}

#Calculate the number of games won by NFC team at home.
table(homewin_nfc$win_conf)

#Number of games won at home by AFC teams whenever they played NFC teams.

homewin_afc<-filter(footballgames, week!= "SuperBowl", footballgames$home_team %in% afc_teams & footballgames$away_team %in% nfc_teams )

dim(homewin_afc)

###To find out which division won when AFC played at home. 
for(i in 1:556){
  if(homewin_afc$winner[i]=="Tie"){
     homewin_afc$win_conf[i]<-"Tied"
  }
  else if(homewin_afc$winner[i]%in% afc_teams){
    homewin_afc$win_conf[i]<-"AFC_Win"
  }
  else{
    homewin_afc$win_conf[i]<-"NFC_Win"  
  }
}
#Calculate the number of games won by AFC team at home.
table(homewin_afc$win_conf)


#Bar plot for the number of games won by NFC team at home. 
plot_nfc_home<-homewin_nfc %>%
  ggplot(aes(x = win_conf, fill = win_conf)) + 
  geom_bar() + 
  geom_text(stat='count', aes(label=..count..), vjust=0) + 
  ggtitle("Number of games won by each conference's team when NFC played at home") +   xlab("Winning Conference") +ylab("Number of games won")+guides(fill=guide_legend(title="Win Conf."))+
    scale_fill_discrete(labels=c('AFC Win', 'NFC Win'))

#Bar plot for the number of games won by AFC team at home. 
plot_afc_home<-homewin_afc %>%
  ggplot(aes(x = win_conf, fill = win_conf)) + 
  geom_bar() + 
  geom_text(stat='count', aes(label=..count..), vjust=0) + 
  ggtitle("Number of games won by each conference's team when AFC played at home") +
  xlab("Winning Conference") +ylab("Number of games won")+guides(fill=guide_legend(title="Win Conf."))
    scale_fill_discrete(labels=c('AFC Win', 'NFC Win', 'Tied')) 

grid.arrange(plot_nfc_home,plot_afc_home, ncol=2)


plot_scatter_nfc_home<- homewin_nfc %>% ggplot(aes(x=yds_win, y= pts_win, color= win_conf))+ geom_point()
plot_scatter_nfc_home

plot_scatter_afc_home<- homewin_afc %>% ggplot(aes(x=yds_win, y= pts_win, color= win_conf))+ geom_point()
plot_scatter_afc_home

```

#### Comparison of points for AFC and NFC teams playing each other.

Here, the objective is to compare the distribution of points won (AFC team vs NFC team) playing in their respective homes. In order to do this, first, dataset that contains only games played at home of NFC teams and away for AFC teams (using 'homewin_nfc' data) is used. The Superbowl games and the games that are 'tied' are not included in the analyses. Then, a box plot of the variable 'pts_win' is created based on 'win_conf' for comparison. Similar analyses is carried out for games played at home of AFC teams and away for NFC teams (using 'homewin_afc' data). Lastly, the two separate plots created are displayed in the same panel for side-by-side comparison.

```{r,echo = FALSE, warning = FALSE}
plot_pts_nfc_home<-ggplot(data=homewin_nfc, aes(x=win_conf,y=pts_win, fill= win_conf))+geom_boxplot() +stat_summary(fun.y=mean, geom="point", shape=20, size=3, color="purple", fill="purple")+xlab("Winning Conference")+ylab("Points Won")+scale_y_continuous(limits = c(0, 60))+ggtitle("Distribution of points won by AFC Vs NFC teams at NFC's home.")+ guides(fill=guide_legend(title="Win Conf."))

ggplotly(plot_pts_nfc_home)

plot_pts_afc_home<-filter(homewin_afc, winner != "Tie") %>% ggplot(aes(x=win_conf, y=pts_win, fill= win_conf))+ geom_boxplot()+stat_summary(fun.y=mean, geom="point", shape=20, size=3, color="purple", fill="purple")+xlab("Winning Conference")+ylab("Points Won")+scale_y_continuous(limits = c(0, 60))+ggtitle("Points won by AFC Vs NFC teams at AFC's home.")+ guides(fill=guide_legend(title="Win Conf."))


ggplotly(plot_pts_afc_home)

grid.arrange(plot_pts_nfc_home, plot_pts_afc_home ,ncol=2)

```

#### Who was winning the most number of SuperBowl from 2000 to 2019? AFC or NFC?

SuperBowl's games are played between two best teams from each conferences:AFC and NFC. Here, the objective is to see the number of Superbowls won by each conference between 2000 to 2019. For this, a subset of original data where 'week' is 'Superbowl' only is filtered out. Then, the winning conference on each of these games is identified with reference from the list 'afc_teams' and stored in a new variable 'win_conf'. Finally, a bar plot is created for the number of Superbowls won by each conference.

```{r}
#Who are winning the most number of SuperBowl from 2000 to 2019?

Superbowl_week<- filter(footballgames, week =="SuperBowl")
dim(Superbowl_week)

###To find out which division won in Superbowl. 
for(i in 1:20){
  if(Superbowl_week$winner[i] %in% afc_teams){
    Superbowl_week$win_conf[i]="AFC Team"
  }
  else{
    Superbowl_week$win_conf[i]="NFC Team"
  }
}  
#Calculate the number of Superbowls Won by AFC and NFC team.
table(Superbowl_week$win_conf)

#Barplot of Superbowl winners.
plot_Superbowl_win_conf<-Superbowl_week %>%
  ggplot(aes(x = win_conf, fill = win_conf)) + 
  geom_bar() + 
  geom_text(stat='count', aes(label=..count..), vjust=0) + 
  ggtitle("Number of games won by each conference in Superbowl") +
  xlab("Winning Conference") +ylab("Number of games won")+guides(fill=guide_legend(title="Win Conf"))
    scale_fill_discrete(labels=c('AFC Win', 'NFC Win')) 

plot_Superbowl_win_conf 
```

#Conclusion:
Objective 1:AFC or NFC, who is winning the most? 
AFC teams has won more number of matches when playing against NFC teams either home or away with only 4 of them being tied. There is not much difference in the distribution of points won between the two conferences in these matches based on the summary statistics.

Objective 2:What does the number of wins at home looks like for AFC vs NFC teams?
Among the games played at NFC's home, NFC won more games compared to AFC. Similarly, AFC teams were winning more games against NFC whenever AFC's teams were playing in their home. 

Objective 3:Comparison of points for AFC and NFC teams playing each other.

There is not much difference in the distribution of points won by winning conferences playing at both NFC's teams and AFC's teams home. 

Objective 4:Who was winning the most number of SuperBowl from 2000 to 2019? AFC or NFC?

Between 2000 and 2019, AFC's teams have won 6 more Superbowl's trophy compared to NFC's teams.     



