---
title: "Project Feedback"
author: "Andrew Ker, Bipin Poudel, Paulus Hermanto"
format: html
editor: visual
---

## Introduction

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load in the data
footballgames <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-04/games.csv')
```

### Data

The data was originally gathered by Pro Football Reference. The data was stored in NFL Scores Main Page under All Game Scores ([data](https://www.pro-football-reference.com/boxscores/game-scores.htm)). The data was used for a tidytuesday project in 2020 ([data](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-02-04/readme.md#gamescsv)) that primarily focused on NFL attendance. The data used for this report was found using the tidytuesday link.

This data originally contained 19 variables related to NFL game scores and outcomes. Out of these 19 variables, 11 were character variables, seven were numeric variables, and one was a time variable. In order to meet project specifications, a categorical variable was needed. To derive a categorical variable, a loop was utilized to replace values of "NA" with a string "0" in the tie column and replace team names with a string "1". This also helped to clean the data, since it does not make sense to list a team winner for a game that ended with a tie. After this loop, the strings of "0" and "1" were converted to a categorical variable. This process essentially created a categorical variable with a value of 0 if an NFL game ended with one team winning and a value of 1 if an NFL game ended in a tie. The code for deriving this variable is shown below.

```{r, echo = TRUE, results = 'hide'}
##### derive categorical variable #####
# replace NA's in tie column with string 0
footballgames$tie[is.na(footballgames$tie)] <- "0"
# replace ties with string 1 instead of string with team name
for(i in 1:5324){
  if(footballgames$tie[i] != "0"){
    footballgames$tie[i] <- "1"
    i < i+1
  }
  else{
    i <- i+1
  }
}

# treat tie as categorical
footballgames$tie <- as.factor(footballgames$tie)


# get structure to check if tie is not a factor
str(footballgames)

footballgames
```

After deriving the categorical variable, additional data cleaning was needed. The "winner" variable contains the name of the winning team; however, this still contained the name of the "winning" team even if the game ended in a tie. To clean this, another loop was utilized to replace the winning team's name with "tie" in the "winner" column, if the game ended in a tie. The code for this is shown below.

```{r, echo = TRUE, results = 'hide'}
# replace ties with string 1 instead of string with team name
for(i in 1:5324){
  if(footballgames$tie[i] == 1){
    footballgames$winner[i] <- "Tie"
    i < i+1
  }
  else{
    i <- i+1
  }
}

footballgames
```

Once the data was cleaned, it still had 19 variables, but with 10 character variables, seven numeric variables, one time variable, and one categorical variable. For more information about the data, please refer to the table below, which contains a list of variables, their descriptions, and their types.

| Variable Name  | Variable Description                                | Variable Type |
|----------------|-----------------------------------------------------|---------------|
| year           | The year in which the game was played               | numeric       |
| week           | The week of the season in which the game was played | character     |
| home_team      | name of the team playing at their home stadium      | character     |
| away_team      | name of the team playing at visiting stadium        | character     |
| winner         | name of the winning team                            | character     |
| tie            | indicates whether the game ended in a tie or not    | categorical   |
| day            | day of the week the game was played on              | character     |
| date           | date the game was played on                         | character     |
| time           | time of the day when the game started               | time          |
| pts_win        | points scored by the winning team                   | numeric       |
| pts_loss       | points scored by the losing team                    | numeric       |
| yds_win        | yards gained by the winning team                    | numeric       |
| turnovers_win  | turnovers committed by the winning team             | numeric       |
| yds_loss       | yards gained by the losing team                     | numeric       |
| turnovers_loss | turnovers committed by the losing team              | numeric       |
| home_team_name | name of the team playing at their home stadium      | character     |
| home_team_city | name of the home team's city                        | character     |
| away_team_name | name of the team playing at visiting stadium        | character     |
| away_team_city | name of the away team's city                        | character     |

## Exploratory Data Analysis

structure of the data

summary statistics

plots of summary statistics

## Analysis

Each of us plans on contributing our findings/analysis here

```{r}
library(tidyverse)
library(magrittr)
library(dplyr)
library(ggplot2)
library(plotly)

# Interactive line plots for the number of wins for selected teams
a=footballgames %>% group_by(year) %>%  count(winner)
b = subset(a, winner== "Baltimore Ravens"|winner=="Chicago Bears"|winner=="Denver Broncos"|winner=="Miami Dolphins"|winner=="New England Patriots")

plot1<- ggplot(data= b, aes(year, n))+geom_line(aes(color=winner))+ xlab("Year")+ylab("Number of wins")+ggtitle("Number of games won by selected teams each year")

ggplotly(plot1)
```

```{r}
#Interactive scatterplot of points won vs yards won for selected teams
m= subset(footballgames, winner== "Baltimore Ravens"|winner=="Chicago Bears"|winner=="Denver Broncos"|winner=="Miami Dolphins"|winner=="New England Patriots")

plot2<-ggplot(data= m)+ geom_point(aes(pts_win, yds_win,color= winner)) +xlab("Points Won")+ ylab("Yards Win")

ggplotly(plot2)
```

Other Ideas for Analysis:

Compare number of games won by the home team to number of games won by the away team Derive winning percentage variable for teams, can compare winning percentage to number of super bowls won compare games won by Broncos to average number of games won for all teams

#### Is there an Advantage to Playing at Home?

It's a common belief that playing any sport (including football) at home stadiums gives the home team an advantage over the visiting team. The results below aim to assess this belief. First, the number of games won by the home team and the number of games won by the road team are tallied. Based off this tally, there were 3,045 NFL games won by the home team from 2000-2019. Over this same stretch of time, there were 2,279 NFL games won by the away team. Clearly, the home team wins more often than the road team, but it is helpful to provide additional context. Since there were 5,324 games played, the home team has won 3,045/5,324 games for a percentage of 57%. This suggests that the home team wins NFL games about 57% of the time. It must be noted however, that these counts are not exact since the data lists a home team, even for games played at a neutral site.

```{r, echo = FALSE, warning=FALSE}
### Comparing the number of games won by the home team to the number of games won by the away team ###

# create a vector of 0's
locationofwin <- rep(c(0), 5324)

# add vector of 0's to the data 
footballgames <- cbind(footballgames, locationofwin)

# for loop to assign values of 1 if the home team won the game
# values of 0 if the home team did not win the game
for(i in 1:5324){
  if(footballgames$winner[i] == footballgames$home_team[i]){
    footballgames$locationofwin[i] <- 1
    i < i+1
  }
  else{
    i <- i+1
  }
}

# count the number of wins by the home team and the number of losses/ties by the home team
table(footballgames$locationofwin)
# 3045
# 2279

# convert locationofwin to categorical
footballgames$locationofwin <- as.factor(footballgames$locationofwin)

# bar plot of games won by home team and games won by away team
footballgames %>%
  ggplot(aes(x = locationofwin, fill = locationofwin)) + 
  geom_bar() + 
  geom_text(stat='count', aes(label=..count..), vjust=0) + 
  ggtitle("Comparing Games Won by Away Teams to Games Won by the Home Team") +
    scale_fill_discrete(labels=c('Away Team Won', 'Home Team Won')) 

```

#### 

#### Which AFC West teams will do well in the next season?

Fans and analysts are often trying to predict which games will be won and lost by their favorite teams. As we have seen from the previous sections, NFL teams are divided into two conferences: AFC and NFC. Since one of our team members is a fan of the Denver Broncos (an AFC West Team), combined with the fact that some AFC West division teams have played the Superbowl 4 times over the past decade (source: [link](https://en.wikipedia.org/wiki/List_of_Super_Bowl_champions); we decided to analyze the pattern of percentage of wins for all teams of AFC West division and plot them as a way to check if there is a trend. By looking at the trend over the past 20 years, we hope that people can assess the teams' past performance and try to predict future performances of AFC West teams.

\*INSERT PLOTS HERE MAYBE?\*

If we were to use 30% as our cutoff percentage to determine whether the year is a bad year, then we see that all four teams have at least one "bad year" where they won less than 30% of the games in each year, throughout 2000-2019. For the Broncos, 2010 was their worst performing year, with 25% games won within that year. However, they win at least 30% of their games for the other nineteen years. For the Chargers, they had gone through three bad years, where they won less than 30% of their games in 2000, 2003, and 2015. For the Raiders, they have had 7 "bad" years over the course of 2000-2019; while the Chiefs had 4 individual years where they won less than 30% of their games (2007, 2008, 2009 and 2012).

Moving on to the "highs", we will look at years in which the teams win at least 60% of the games throughout 2000-2019. From this point on, the focus will be to dissect each team's performance over the course of 20 years. The Broncos had a good year in 2000, where they won more around 65% of their games, then started performing pretty decently (within 50%-60% games won) from 2001-2004. They won approximately 78% of the games they played in 2005, and even won the Divisional Round. However, it appears that their performance sharply declined within the next 5 years, leading up to their lowest year in 2010 where they won only 25% of the games in that year. They successfully bounced back and had an impressive trajectory over the next 5 years, starting in 2011 when they won 50% of their games, then to consistently, over the next four years, won at least 70% of their games. In 2013 and 2015, the Broncos managed to pull off victories in approximately 79% of their games; which are also signified by the team moving on to play the Superbowl in both years. After winning the Superbowl in 2015, unfortunately, they did not have any good years from 2016-2019.

The Chargers started from the bottom at the year of 2000, when they won less than 10% of their games (they only won one game). The team did not have any memorable years until after 2003. Over the next 10 years, the trajectory of their wins to start toggling between average, good and even exceptional. Starting in 2004, where they had a noticeable comeback with 70% of games won in that year. While the following year, 2005, might be considered to be unremarkably average to many (\~56% wins), the team was able to improve exceptionally in 2006 by winning more than 80% of their games. The following years, the Chargers kept having "average" years (if we are only looking at winning percentages) with the exception of 2007 and 2009. Surprisingly, in 2015, the team's performance plummeted when they won less than 30% of games that year. They managed to come back stronger over the next 3 years, leading up to another impressive year in 2018 where they won approximately 72% of their games; but plummeted again the following year, in 2019, when they only won approximately 30% of the time. All in all, while the Chargers never made it to the Superbowl; they had some pretty impressive years in 2006, 2009, 2018, where they made the Divisional round, and even won more than 70% of their games.

For the first 13 years, the Chiefs only had two years in which we consider to be good years: 2003 and 2005. In both of those years, they managed to pull off 76% of wins and 63% of wins. Other than those triumphant years, the team was either performing moderately, or extremely terribly. Although 2007, 2008, 2009 and 2012 were the terrible years for the Kansas-based team, they managed to turn it around starting from year 2013 where they consistently won more games than lost. In fact, the year 2012 can almost be seen as a turning point for the team, in which we start seeing a slight upward trend for the Chiefs' winning percentages. More impressively, the Chiefs even managed to win more than 70% of games in 2016, 2018, 2019. While our data only shows games up to 2019, in reality, the team had a very strong track record throughout 2020-2022 as well (source: <https://www.espn.com/nfl/standings>) which some might have predicted from the successful streak in those previous years.

Lastly, the Raiders possess what could be seen as the least impressive track record out of the teams discussed so far. The team started off strong from the year 2000-2002, where they won more than 60% of the games within those years. After 2002, their winning percentages dropped significantly to below 30% in 2003. The team did not significantly improve over the next 6 years. A somewhat noteworthy improvement was made in 2010 and 2011 where they won 50% of games. However, the percentages plummeted again after unsatisfactory years: 2012, 2013, and 2014; as the team won below 30% of their games. The team bounced back in 2015 and 2016; outstandingly, the team was able to reach 70% of wins in 2016. The achievement, nevertheless, failed to last as their percentages plummeted again after the high in 2016 and failed to bounce back until the year of 2019.

Overall, we realize that it is difficult to predict wins based on trends of winning percentages, as there are other factors in place that are affecting team performances such as: players getting traded (roster changes), coaching styles (some coaches are more oriented towards offense/defense), game schedules, weather conditions, etc. Furthermore, every team's performance on the AFC West Division has been extremely volatile, making it more difficult for their wins to be predictable and to observe stability of their performance in future years. However, one team that stuck out the most from our observation is the Chiefs. The plot above shows that the team has maintained high winning percentages over the past 6 years, which could have potentially been able to predict their win for the 2020 Superbowl (source: <https://en.wikipedia.org/wiki/List_of_Super_Bowl_champions>).





## Conclusion

## Future Ideas

## References
