---
title: "Stat 850 Project Description"
author: "Andrew Ker, Bipin Poudel, Paul Hermanto"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(digits = 3)
```

```{r}
##### Load in the Dataset #####
footballgames <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-04/games.csv')
```



## Instructions

Each member of your team should modify this document in some way and push their modifications to the repository in a separate commit. This will ensure that you have set your repository up in a way that ensures all group members are working with the same repository. 

Note that you must compile your readme (this document) for it to properly display as part of your github repository. 

Once you have received feedback on your project proposal (via Canvas) you may alter this README so that it describes your final project instead of the project proposal. 

## Data Set
Provide a link to your dataset here, and a description of the variables in the data set.

Link:https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-04/games.csv


## Potential Topics to Explore Using the Data Set
Describe avenues you might explore using the data
-Paul: We can try to observe/filter out the the wins of a certain team. We can try to calculate average points won. We can see if there's any tie.
-Bipin: Prop 1: A line plot of total points won by each team through out the years.
-Bipin: Prop 2: The total number of games won by Denver Broncos through out the years from 2000 to 2019. 
-


## Group Members
List all of the project contributors here.

Andrew Ker
Bipin Poudel
Paulus Hermanto

